---
title: "Setup and Installation"
---

# Install the packages the Hudson way

Most Hudson packages live in GitLab, so the first hurdle is not R code. It is
package access and installation discipline.

## Step 1: create a GitLab personal access token

At a high level:

1. Open your GitLab account settings.
2. Go to the access token area.
3. Create a token with the scopes needed to read packages and repositories.
4. Store it safely.

Then add it to `~/.Renviron`:

```{r}
#| eval: false
usethis::edit_r_environ()
```

Add a line like:

```text
GITLAB_PAT=glpat-your-token-goes-here
```

Restart R after saving.

::: {.callout-warning}
Do not hard-code tokens in scripts, notebooks, or project files.
:::

## Step 2: install `hiverse`

`hiverse` is the meta-package that gives you a convenient entry point.

```{r}
#| eval: false
install.packages("devtools")
devtools::install_gitlab("isazi/hudson-packages/hiverse@main")
library(hiverse)
```

## Step 3: install the common package set

`hiverse` can install the common packages for you:

```{r}
#| eval: false
hiverse::install_hudson_packages("common")
```

That common set currently includes the main packages new employees are most
likely to use first, such as `hiconnect`, `himunge`, `hiproc`, `hidate`,
`hilog`, `higpt`, `higrate`, `hidrogen`, and `himodel`.

## Step 4: install optional packages only when needed

Some packages are more specialist and should usually be installed when the work
demands them:

```{r}
#| eval: false
devtools::install_gitlab("isazi/hudson-packages/hipoint")
devtools::install_gitlab("isazi/hudson-packages/hisched")
```

Other optional packages exist as well, especially for supply-chain work, but
you do not need all of them on day one.

## A quick installation check

Once packages are installed, run a quick sanity check:

```{r}
#| eval: false
required_pkgs <- c(
  "hiverse", "himunge", "hiconnect", "hidate", "hilog", "hiproc",
  "higpt", "higrate", "hidrogen", "himodel"
)

data.frame(
  package = required_pkgs,
  installed = vapply(required_pkgs, requireNamespace, logical(1), quietly = TRUE)
)
```

## Recommended learning order

If you are a new joiner and want the shortest path to competence:

1. `himunge`
2. `hiconnect`
3. `hidate`
4. `hilog`
5. `hiproc` if you touch processes
6. `higpt`
7. `higrate`
8. `hidrogen`
9. `himodel`
10. `hipoint` or `hisched` if your project calls for them

## Try it

1. Install `hiverse`.
2. Install the common packages.
3. Verify that `requireNamespace("himunge", quietly = TRUE)` returns `TRUE`.
4. Write down which packages failed to install, if any, and why.

## Stretch

Create a small script called `check_hiverse_install.R` that:

- checks a package list
- prints which packages are installed
- prints `sessionInfo()` at the end

This is a good first "works on my machine" diagnostic to keep around.

## Think

Why is `hiverse` useful even if you could install every package directly with
`devtools::install_gitlab()`?
