---
title: "Core Data Packages"
---

# The packages you will use most often

For most data scientists at Isazi, the center of gravity is:

- `himunge` for understanding and cleaning data
- `hiconnect` for getting data in and out of Hudson databases
- `hidate` for calendar logic that appears in real business work
- `hilog` for making scripts and processes observable from the start

If you become fluent with these four, the rest of the hiverse becomes easier.

## `himunge`: your first stop for unfamiliar data

`himunge` is the strongest all-round package in the ecosystem. It is useful
because it helps with the actual first questions people ask about a dataset:

- what do these columns mean?
- what looks like a key?
- which columns determine which others?
- where is the data inconsistent?

### A tiny messy dataset

The following chunk creates a toy order table with a hidden key problem:

```{r}
orders_dirty <- data.frame(
  customer_id = c("C001", "C001", "C002", "C002", "C003", "C003"),
  order_id = c("O100", "O100", "O200", "O200", "O300", "O300"),
  order_date = c("2025-01-03", "2025-01-03", "2025-01-10", "2025-01-10", "2025-02-01", "2025-02-01"),
  order_status = c("shipped", "cancelled", "processing", "processing", "shipped", "shipped"),
  amount = c("1,200", "1,200", "850", "850", "2,000", "2,000"),
  stringsAsFactors = FALSE
)

dup_counts <- aggregate(
  rep(1, nrow(orders_dirty)),
  by = list(order_id = orders_dirty$order_id),
  FUN = length
)

barplot(
  dup_counts$x,
  names.arg = dup_counts$order_id,
  col = "#0b6e69",
  main = "Rows per order_id in the toy dataset",
  ylab = "Count"
)
abline(h = 1, lty = 2, col = "#a44a1f")
```

The point is not the bar chart itself. The point is that duplicate-looking keys
are one of the fastest ways to discover real data problems.

### Example workflow

```{r}
#| eval: false
library(himunge)

clean_orders <- orders_dirty |>
  himunge::fix_column_names() |>
  himunge::convert_hidden_dates() |>
  himunge::convert_hidden_numerics()

himunge::find_primary_keys(clean_orders)

viol <- himunge::find_indices_violating_key(
  clean_orders,
  candidate_key = c("customer_id", "order_id"),
  make_reduced_dataframe = TRUE
)

viol$rows
viol$cols
viol$data
```

### Why this matters

This is the kind of workflow that helps a data scientist answer:

- "Is this really one row per order?"
- "Which columns disagree when the key stays fixed?"
- "Do I trust this table enough to model from it?"

### Try it

1. Add a `sales_rep` column that is perfectly determined by `customer_id`.
2. Run `himunge::get_column_dependency_tree()` on the cleaned data.
3. Explain which columns appear downstream and why.

### Stretch

Build a new toy admissions dataset with columns like:

- `student_id`
- `campus`
- `programme`
- `intake_month`
- `fee_band`
- `advisor`

Then:

1. use `find_primary_keys()`
2. introduce a deliberate key violation
3. use `find_indices_violating_key()` to isolate the bad rows
4. write a one-paragraph summary of the issue as if you were reporting it to a client

## `hiconnect`: the main route into Hudson data

`hiconnect` matters because most real Hudson work eventually touches a database.
It is not just a thin wrapper over `DBI`; it encodes Hudson conventions around
schemas, outputs, and process configuration.

### The important idea

Before you can query anything, you usually need a valid process config file.
The package README calls this out explicitly.

```{r}
#| eval: false
library(hiconnect)

generate_first_hiconnect_config(
  host = "your-host",
  user = "your-user",
  password = "your-password"
)
```

Then the common workflow is:

```{r}
#| eval: false
con <- hiconnect::pg_connect(c("scratch", "process"))
on.exit(DBI::dbDisconnect(con), add = TRUE)

row_count <- hiconnect::pg_count("scratch", "some_table", con = con)
sample_tbl <- hiconnect::pg_get_table("scratch", "some_table", limit = 100, con = con)
```

### How to think about `hiconnect`

Use it when you need to:

- inspect tables quickly
- read or write Hudson data safely
- build process-aware code
- avoid hand-written connection boilerplate in every script

### Try it

This exercise needs Hudson database access.

1. Connect to one schema you are allowed to access.
2. Pick one table and count its rows.
3. Pull the first few rows.
4. Identify what you believe the primary key might be before using `himunge`.

### Stretch

Use `hiconnect` and `himunge` together:

1. read a table from Hudson
2. propose a key with `find_primary_keys()`
3. check whether any columns violate that key
4. write down whether the table looks model-ready

## `hilog`: log early, log clearly

`hilog` is small, but it is worth learning early because it encourages a good
habit: code should say what it is doing while it runs.

That matters in:

- long-running scripts
- data-quality audits
- package code
- Hudson processes
- debugging interactive work that later becomes production code

### The core idea

The logging functions support `sprintf`-style interpolation directly through
their `...` arguments, so you can write concise, structured messages without
building strings manually.

```{r}
#| eval: false
library(hilog)

hilog::set_level_info()

hilog::info("Loaded %i rows from %s", 1250, "orders_extract.csv")
hilog::info("Coverage after cleaning: %.1f%%", 97.4)
```

That style is worth copying. It keeps the message template stable and pushes
the changing values into arguments.

A typical emitted line looks like:

```text
[2026-03-16 10:15:00 audit_orders] INFO: Starting order audit on 6 rows and 5 columns
```

The timestamp and calling context come for free, which is why this is more
useful than scattering `print()` calls through your code.

### A more realistic example

```{r}
#| eval: false
audit_orders <- function(df) {
  hilog::info("Starting order audit on %i rows and %i columns", nrow(df), ncol(df))

  key_sets <- himunge::find_primary_keys(df)
  hilog::info("Found %i candidate key sets", length(key_sets))

  if (length(key_sets) > 0) {
    hilog::info("Smallest candidate key: %s", paste(key_sets[[1]], collapse = ", "))
  } else {
    hilog::error("No primary key candidate found for current dataset")
  }
}
```

In practice, that means you do not need to guess where a script got to before
it failed or started behaving strangely.

### Debug level

```{r}
#| eval: false
hilog::set_level_debug()
hilog::debug("Duplicate order_id count: %i", sum(duplicated(orders_dirty$order_id)))
```

Use `info()` for progress, `debug()` for extra detail, and `error()` for
messages that should always be visible.

### Try it

1. Add `hilog::info()` calls to one of your own analysis scripts.
2. Log:
   - how many rows were loaded
   - how many columns were selected
   - what key you think the table has
3. Rewrite one message using `%i`, `%s`, or `%.2f` formatting instead of `paste()`.

### Stretch

Take the earlier `himunge` toy exercise and add logging at each step:

1. raw row count
2. cleaned row count
3. number of candidate key sets found
4. whether violations were detected

Then imagine that script runs overnight. Which log lines would help you most
the next morning?

## `hidate`: small package, very high practical value

Many Isazi projects are date-heavy. Forecast cutoffs, planning windows, working
days, weekly rollups, and month boundaries show up constantly.

### Simple example

```{r}
#| eval: false
library(hidate)

date <- as.Date("2026-03-16")

hidate::start_of_month(date)
hidate::end_of_month(date)
hidate::start_of_week(date)
hidate::end_of_week(date)
hidate::diff_weekdays(c(as.Date("2026-03-16"), as.Date("2026-03-31")))
```

### A visual: working-day accumulation

```{r}
march_days <- seq(as.Date("2026-03-01"), as.Date("2026-03-31"), by = "day")
weekday_flag <- weekdays(march_days) %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
cum_weekdays <- cumsum(weekday_flag)

plot(
  march_days,
  cum_weekdays,
  type = "l",
  lwd = 3,
  col = "#0b6e69",
  xlab = "Date",
  ylab = "Cumulative weekdays",
  main = "Why date helpers matter in planning work"
)
grid()
```

### Try it

1. Pick a date relevant to your project.
2. Compute its month start, month end, week start, and week end.
3. Work out how many weekdays remain in that month.

### Stretch

Write a function called `planning_window()` that returns:

- the current week start
- the current month end
- the number of weekdays between them

Then compare your result with `hidate::diff_weekdays()`.

## Think

Which is riskier on a real project:

- getting the model wrong
- or getting the date logic wrong before the model is even run?
