---
title: "Platform Workflows"
---

# Packages for Hudson process work

Not every analyst needs these every day. But if you work close to the Hudson
platform, they matter a lot.

## `hiproc`: understanding processes, queues, and dependencies

`hiproc` is important because Hudson is not just a collection of tables. It is
also a process system. When something fails, runs late, or produces a strange
output, you often need process context, not just data context.

### What `hiproc` is good for

- summarising process events
- looking up dependencies
- inspecting files and run parameters
- understanding what happened before and after a process run

### Example

The book renders this example against a tiny mocked process registry so that
the actual `hiproc` calls can run without Hudson access.

```{r}
#| echo: false
with_mocked_hiconnect_pg_get_query <- function(mock_fn, expr) {
  ns <- asNamespace("hiconnect")
  old <- get("pg_get_query", envir = ns)
  unlockBinding("pg_get_query", ns)
  assign("pg_get_query", mock_fn, envir = ns)
  lockBinding("pg_get_query", ns)
  on.exit({
    unlockBinding("pg_get_query", ns)
    assign("pg_get_query", old, envir = ns)
    lockBinding("pg_get_query", ns)
  }, add = TRUE)
  force(expr)
}

mock_process_pg_get_query <- function(schemas = NULL, statement = NULL, params = NULL, con = NULL, ...) {
  sql <- gsub("\\s+", " ", statement)

  if (grepl("from process.process pp left join process.event pe", sql, fixed = TRUE)) {
    return(data.frame(
      name = c("ingest_orders", "ingest_orders", "build_dashboard"),
      process_id = c(1L, 1L, 2L),
      event_type = c("Success", "Failed", "Success"),
      first_run_date = as.Date(c("2026-01-01", "2026-01-05", "2026-01-02")),
      last_run_date = as.Date(c("2026-01-10", "2026-01-12", "2026-01-15")),
      last_process_result_id = c(10L, 12L, 15L),
      num_runs = c(5L, 1L, 4L)
    ))
  }

  if (grepl("select process_id from process.process where name = \\$1", sql)) {
    ids <- c(ingest_orders = 1L, build_dashboard = 2L)
    return(data.frame(process_id = ids[[params[[1]]]]))
  }

  if (grepl("process_dependency_process", sql)) {
    return(data.frame(
      name = c("build_dashboard"),
      process_id = c(2L),
      dependency_process_id = c(1L),
      dependency_process_name = c("ingest_orders"),
      output_schema = c("scratch"),
      output_table = c("orders"),
      group_id = c(1L),
      group_name = c("upstream process"),
      data_only = c(FALSE),
      cyclic = c(FALSE)
    ))
  }

  if (grepl("process_dependency_static", sql)) {
    return(data.frame(
      process_id = c(1L),
      static_table_id = c(101L),
      static_table_name = c("calendar_dim"),
      static_table_schema = c("static"),
      static_table_table = c("calendar_dim"),
      group_id = c(11L),
      group_name = c("calendar"),
      data_only = c(TRUE)
    ))
  }

  if (grepl("process_dependency_scheduled", sql)) {
    return(data.frame(
      process_id = c(1L),
      cron_min = c("0"),
      cron_hour = c("6"),
      cron_day = c("*"),
      cron_month = c("*"),
      cron_week_day = c("1-5"),
      group_id = c(12L),
      group_name = c("weekday trigger"),
      data_only = c(FALSE)
    ))
  }

  if (grepl("process_dependency_manual", sql)) {
    return(data.frame(
      process_id = integer(),
      manual_dependency = character(),
      group_id = integer(),
      group_name = character(),
      data_only = logical()
    ))
  }

  stop("Unhandled SQL in mock: ", sql)
}
```

```{r}
library(hiproc)

offline_process_view <- with_mocked_hiconnect_pg_get_query(mock_process_pg_get_query, {
  list(
    events = hiproc::event_summary(),
    deps = hiproc::get_dependencies(
      "build_dashboard",
      recursive = TRUE,
      con = structure(list(), class = "mock_con")
    )
  )
})

offline_process_view$events[, c("process_name", "last_status", "diagnostic")]
offline_process_view$deps$process[, c("name", "dependency_process_name", "depth")]
```

### Process thinking

```{mermaid}
flowchart TD
  A["Raw input file arrives"] --> B["Ingestion process runs"]
  B --> C["Table is updated"]
  C --> D["Downstream process runs"]
  D --> E["Dashboard or forecast changes"]
  F["hiproc"] --> B
  F --> D
```

When work breaks in production, this is the package that helps you move from
"the number looks wrong" to "which process or dependency likely caused it?"

### Try it

This exercise needs platform access.

1. Run `event_summary()`.
2. Pick one process that looks active or familiar.
3. Inspect its dependencies.
4. Sketch the upstream/downstream flow in your own words.

### Stretch

Choose one process output table and one upstream dependency. Ask:

- what data enters here?
- what process transforms it?
- what would fail silently vs loudly?

Then write a short incident note as if you were on support duty.

## `higest`: standardised ingestion logic

`higest` is not a day-one analyst package, but it is useful once you start
building repeatable ingestion flows.

The package frames ingestion as four steps:

1. prepare
2. read
3. standardise
4. write

### Example pipeline shape

The full `prepare()` path also needs a process config file and a writable Hudson
target. For a day-one local sandbox, you can start one step later with a tiny
spec and still run the package pipeline end to end.

```{r}
library(higest)

tmp_csv <- tempfile(fileext = ".csv")
write.csv(
  data.frame(order_id = c("O100", "O101"), amount = c(1200, 950)),
  tmp_csv,
  row.names = FALSE
)

my_format_df <- data.frame(
  orig_name = c("order_id", "amount"),
  clean_name = c("order_id", "amount"),
  final_name = c("order_id", "amount"),
  format = c("character", "numeric"),
  stringsAsFactors = FALSE
)

pipeline <- list(
  ingestion_process_name = "toy_orders",
  file = list(path = tmp_csv),
  format_df = my_format_df,
  output_table = list(schema = "scratch", table = "toy_orders")
)
class(pipeline) <- "higest_spec"

pipeline <- pipeline |>
  higest::add_read_spec_csv() |>
  higest::add_transformation_identity()

pipeline$write_spec <- list(type = "collect")
pipeline$write_fn <- function(hs) hs$final_data

invisible(utils::capture.output({
  result <- higest::run_higest_pipeline(pipeline)
}))

result$result
```

The important lesson is not the exact function names. It is the idea that
ingestion should be declared and inspectable, not hidden inside a one-off script.

## Other platform-adjacent helpers

- `hifile`: import/export helpers for files and file tables
- `hidoc`: report and data-dictionary helpers
- `hidash`: dashboard helpers for Shiny-based Hudson work
- `hilog`: simple logging utilities that support package and process code

You do not need to master these immediately, but you should know they exist.

## Think

When a stakeholder says "the dashboard changed unexpectedly", what is the right
first question:

- "Which model did we use?"
- or "Which process ran, with which inputs, and what changed upstream?"
