---
title: "Platform Workflows"
---

# Packages for Hudson process work

Not every analyst needs these every day. But if you work close to the Hudson
platform, they matter a lot.

## `hiproc`: understanding processes, queues, and dependencies

`hiproc` is important because Hudson is not just a collection of tables. It is
also a process system. When something fails, runs late, or produces a strange
output, you often need process context, not just data context.

### What `hiproc` is good for

- summarising process events
- looking up dependencies
- inspecting files and run parameters
- understanding what happened before and after a process run

### Example

```{r}
#| eval: false
library(hiproc)

events <- hiproc::event_summary()
head(events)

deps <- hiproc::get_dependencies("some_process_name")
deps
```

### Process thinking

```{mermaid}
flowchart TD
  A["Raw input file arrives"] --> B["Ingestion process runs"]
  B --> C["Table is updated"]
  C --> D["Downstream process runs"]
  D --> E["Dashboard or forecast changes"]
  F["hiproc"] --> B
  F --> D
```

When work breaks in production, this is the package that helps you move from
"the number looks wrong" to "which process or dependency likely caused it?"

### Try it

This exercise needs platform access.

1. Run `event_summary()`.
2. Pick one process that looks active or familiar.
3. Inspect its dependencies.
4. Sketch the upstream/downstream flow in your own words.

### Stretch

Choose one process output table and one upstream dependency. Ask:

- what data enters here?
- what process transforms it?
- what would fail silently vs loudly?

Then write a short incident note as if you were on support duty.

## `higest`: standardised ingestion logic

`higest` is not a day-one analyst package, but it is useful once you start
building repeatable ingestion flows.

The package frames ingestion as four steps:

1. prepare
2. read
3. standardise
4. write

### Example pipeline shape

```{r}
#| eval: false
pipeline <- higest::prepare("process X", my_format_df) |>
  higest::add_read_spec_csv() |>
  higest::add_transformation_identity() |>
  higest::add_write_spec_replace()

higest::run_higest_pipeline(pipeline)
```

The important lesson is not the exact function names. It is the idea that
ingestion should be declared and inspectable, not hidden inside a one-off script.

## Other platform-adjacent helpers

- `hifile`: import/export helpers for files and file tables
- `hidoc`: report and data-dictionary helpers
- `hidash`: dashboard helpers for Shiny-based Hudson work
- `hilog`: simple logging utilities that support package and process code

You do not need to master these immediately, but you should know they exist.

## Think

When a stakeholder says "the dashboard changed unexpectedly", what is the right
first question:

- "Which model did we use?"
- or "Which process ran, with which inputs, and what changed upstream?"
