---
title: "AI-Era Extensions"
---

# The newer AI layer

These packages sit on top of the more traditional Hudson stack.

- `higpt` is the foundation.
- `higrate` is about profiling and knowledge-base generation.
- `hidrogen` is about agentic workflows.

## `higpt`: the foundation layer

If you only learn one AI package well, learn `higpt`.

It is the main LLM interface package for the ecosystem and supports:

- chat
- tool calling
- structured outputs
- embeddings
- image and audio workflows
- multiple providers

### Two useful mental models

- `continue_chat()` and `chat_with_tools()` are the lower-level interfaces.
- `make_model() |> ... |> submit()` is the newer tidy DSL.
- The builder DSL is usually the better teaching surface because it keeps
  prompts, schemas, tools, logging, and async settings in one object.

The OpenAI-based examples below run live when `OPENAI_API_KEY` is available in
your environment. The only exception is the final `mutate_llm()` snippet, which
is shown without execution because it intentionally fans out multiple model
calls.

```{r}
#| echo: false
openai_key_available <- nzchar(Sys.getenv("OPENAI_API_KEY"))
```

::: {.callout-note}
To render the live `higpt` examples, set `OPENAI_API_KEY` in your environment.
`higpt` also supports `GEMINI_API_KEY`, `ANTHROPIC_API_KEY` or
`CLAUDE_API_KEY`, and `XAI_API_KEY` or `GROK_API_KEY`.
:::

```{r}
library(higpt)

capable_models <- subset(
  higpt::get_available_models("continue_chat"),
  structured_outputs | supports_vision,
  select = c(model, api, structured_outputs, supports_vision)
)

utils::head(capable_models, 6)
```

### Model aliases: `-latest` and friends

One of the most practical `higpt` features is that many model names are aliases
rather than hard-coded vendor IDs.

```{r}
alias_models <- subset(
  higpt::get_available_models("continue_chat"),
  model %in% c(
    "gpt-latest",
    "gpt-chat-latest",
    "gpt-mini-latest",
    "claude-sonnet-latest",
    "gemini-pro-latest",
    "grok-latest"
  ),
  select = c(model, api, api_model, category)
)

alias_models
```

As of 2026-03-17 in the current package registry, those aliases resolve to
concrete IDs via the `api_model` column. For example:

- `gpt-latest` currently resolves to `gpt-5.4`
- `gpt-chat-latest` currently resolves to `gpt-5.3-chat-latest`
- `claude-sonnet-latest` currently resolves to `claude-sonnet-4-6`
- `gemini-pro-latest` currently resolves to `gemini-3.1-pro-preview`

This is a good default rule:

- use `*-latest` when you want the newest capabilities without editing code
- use `*-stable` when you want a more conservative default for shared workflows
- use a dated or explicit concrete model ID when you need exact reproducibility

### Builder-first workflow

The tidy DSL is useful when you want a reusable chat template rather than a
one-off prompt.

```{r}
quality_chat <- higpt::make_model("gpt-5-mini") |>
  higpt::add_system_prompt("You are a concise data quality assistant.") |>
  higpt::add_user_message("List three first checks for a new orders table.") |>
  higpt::set_reasoning_effort("minimal") |>
  higpt::set_options(temperature = 0.2)

higpt::preview(quality_chat)
```

```{r, eval=openai_key_available}
builder_result <- quality_chat |>
  higpt::submit()

higpt::collect_text(builder_result)
```

### Structured outputs with schema builders

The `make_schema()` / `add_field()` / `finalize_schema()` pattern is one of the
most useful additions in the newer `higpt` material because it makes it easier
to connect LLM outputs to downstream tables.

```{r}
check_schema <- higpt::make_schema("quality_check") |>
  higpt::add_field("check_name", "string") |>
  higpt::add_field("why_it_matters", "string") |>
  higpt::finalize_schema()
```

```{r, eval=openai_key_available}
schema_result <- higpt::make_model("gpt-mini-latest") |>
  higpt::add_user_message("Name one important validation for an orders table.") |>
  higpt::set_schema(check_schema, multiple_rows = FALSE) |>
  higpt::submit()

schema_result |>
  higpt::collect_df()
```

If a provider ignores native structured outputs, `enforce_schema()` is the
fallback that cleans the text response into the requested shape.

### Tools with the builder

```{r}
add_numbers <- higpt::make_tool("add_numbers", "Add two numbers together") |>
  higpt::add_parameter("a", "number", required = TRUE) |>
  higpt::add_parameter("b", "number", required = TRUE) |>
  higpt::finalize_tool()

handlers <- list(
  add_numbers = function(args) as.character(args$a + args$b)
)
```

```{r, eval=openai_key_available}
tool_result <- higpt::make_model("gpt-mini-latest") |>
  higpt::add_user_message("What is 12 + 30? Use the tool.") |>
  higpt::add_tool(add_numbers) |>
  higpt::set_tool_choice("auto") |>
  higpt::set_tool_handlers(handlers) |>
  higpt::submit()

higpt::collect_text(tool_result)
```

### Prompt files and row-wise workflows

Once prompts become team assets, keep them on disk rather than burying them in
scripts. `higpt` now supports prompt templates, prompt-schema bundles, and
row-wise workflows like `mutate_llm()`.

```{r}
prompt_file <- tempfile(fileext = ".md")
writeLines(
  c(
    "You are supporting the {{team}} team.",
    "Summarise the main risks for {{topic}} in two bullets."
  ),
  prompt_file
)

templated_chat <- higpt::make_model("gpt-5-mini") |>
  higpt::set_prompt_template(prompt_file, role = "user") |>
  higpt::apply_prompt_template(list(team = "operations", topic = "late deliveries"))

higpt::preview(templated_chat)
```

```{r}
risk_schema <- higpt::make_schema("risk_summary") |>
  higpt::add_field("topic", "string") |>
  higpt::add_array_field("actions", "string") |>
  higpt::finalize_schema()

bundle <- higpt::bundle_prompt_schema(prompt_file, risk_schema)

bundle$prompt_path
names(bundle$schema$fields)
```

```{r}
driver_notes <- data.frame(
  metric = c("stockouts", "late deliveries", "returns"),
  stringsAsFactors = FALSE
)
```

```{r}
#| eval: false
notes <- higpt::mutate_llm(
  paste("Write one business interpretation for", driver_notes$metric),
  model_name = "gpt-mini-latest",
  .progress = FALSE
)

data.frame(metric = driver_notes$metric, note = notes)
```

If your team uses local or OpenAI-compatible endpoints, `register_custom_model()`
and `list_endpoint_models()` are worth learning as well because the same builder
patterns then work against internal infrastructure too.

### Try it

This exercise needs an API key.

1. Use `get_available_models()` to find one model with structured outputs and
   one with vision support.
2. Build a `make_model()` template for a data-quality review.
3. Add a schema and extract a one-row tibble with
   `enforce_schema() |> collect_df()`.

### Stretch

Create a small reusable workflow that includes all of these:

- a prompt file on disk
- a schema built with `make_schema()`
- either `mutate_llm()` across several rows or a custom tool built with `make_tool()`
- optional logging with `enable_logging()`

## `higrate`: from data frames to reusable knowledge

`higrate` is one of the most interesting newer packages because it connects
classic data profiling with AI-generated narratives and relationship discovery.

### The core move

Take in-memory tables and turn them into:

- candidate keys
- summary statistics
- relationship suggestions
- reusable knowledge-base entries

### Example with no LLM requirement

```{r}
library(higrate)

orders_dirty <- data.frame(
  record_id = 1:6,
  customer_id = c("C001", "C001", "C003", "C004", "C002", "C002"),
  order_id = c("O100", "O101", "O300", "O400", "O200", "O200"),
  order_date = as.Date(c("2025-01-03", "2025-01-08", "2025-02-01", "2025-02-05", "2025-01-10", "2025-01-10")),
  order_status = c("shipped", "processing", "shipped", "processing", "processing", "cancelled"),
  amount = c(1200, 950, 2000, 500, 850, 850)
)

orders_desc <- higrate::generate_table_description(
  orders_dirty,
  table_name = "orders",
  use_llm = FALSE,
  use_llm_columns = FALSE
)

customers <- data.frame(
  customer_id = c("C001", "C002", "C003"),
  segment = c("retail", "retail", "wholesale"),
  region = c("Gauteng", "Western Cape", "KZN")
)

customers_desc <- higrate::generate_table_description(
  customers,
  table_name = "customers",
  use_llm = FALSE,
  use_llm_columns = FALSE
)

rel <- higrate::generate_relationships(
  orders_desc,
  customers_desc,
  use_llm = FALSE
)

rel$proposed_relationships
```

### Relationship thinking

```{mermaid}
flowchart LR
  A["orders"] -->|"customer_id"| B["customers"]
  A -->|"product_id"| C["products"]
  B --> D["customer narrative"]
  C --> E["product narrative"]
  A --> F["table profile"]
  F --> G["knowledge base entry"]
```

### Try it

1. Build two toy tables with a clear join path.
2. Run `generate_table_description(..., use_llm = FALSE)`.
3. Run `generate_relationships(..., use_llm = FALSE)`.
4. Decide whether the discovered relationship is actually useful.

### Stretch

Choose a real project table pair and design a small knowledge base:

- which tables deserve narratives?
- which relationships matter?
- what would you want a new teammate to learn from the saved KB?

## `hidrogen`: agentic workflows on top of the stack

`hidrogen` is the package to look at when you want orchestrated AI workflows
rather than one-off prompts.

### What to focus on

- `run_data_explorer()` for exploration
- `run_model_explorer()` for model-oriented workflows
- workflow YAML files
- prompt files
- tool registry design

### Example

This example is shown but not executed in the book. `hidrogen` runs through
`higpt`, so it needs the same provider setup as the live `higpt` examples or a
registered local/OpenAI-compatible model.

```{r}
#| eval: false
library(hidrogen)

exploration <- hidrogen::run_data_explorer(
  mtcars,
  sample_rows = 3,
  model_name = "gpt-mini-latest"
)

exploration$report$executive_summary
exploration$report$key_insights
```

### Why it matters

`hidrogen` teaches a more important lesson than any one workflow: if you want
AI systems to be dependable, they need structure, tools, and auditable steps.

### Try it

1. Read one prompt file from `hidrogen`.
2. Identify which parts are domain knowledge and which parts are workflow control.
3. Suggest one improvement to make the workflow safer or clearer.

### Stretch

Design a new workflow idea for your team. Example:

- data dictionary builder
- process incident explainer
- first-pass forecast QA agent

Write:

1. the goal
2. the required tools
3. the main failure modes

## Think

Which of these creates more value in practice:

- an LLM that can talk about your data
- or a workflow that turns data knowledge into reusable team memory?
