---
title: "Modeling and Forecasting"
---

# Where `himodel` fits

`himodel` is not the most universal package in the hiverse, but it is important
for forecasting and structured modeling work.

It is best understood as a modeling utility package with a strong forecasting
flavor, not as a general replacement for all of R's modeling ecosystem.

## What it is useful for

- forecast error calculation
- forecast reshaping and reallocation
- helper utilities around time-based model workflows

## A minimal example

```{r}
#| eval: false
library(himodel)

forecast_eval <- data.frame(
  sku = c("A", "A", "A", "B", "B", "B"),
  actual = c(100, 120, 110, 80, 75, 90),
  forecast = c(95, 130, 100, 82, 70, 88)
)

himodel::compute_error(
  forecast_eval,
  measure_level = c("sku"),
  report_level = c("sku"),
  actual_var = "actual",
  pred_var = "forecast"
)
```

## Why this matters

A lot of poor forecasting practice starts before the model itself:

- wrong aggregation level
- poor error reporting
- hidden bias
- weak hierarchy handling

`himodel` is useful when you want a standard way to talk about those things.

## A simple visual

```{r}
actual <- c(100, 120, 110, 80, 75, 90)
forecast <- c(95, 130, 100, 82, 70, 88)
period <- seq_along(actual)

matplot(
  period,
  cbind(actual, forecast),
  type = "b",
  pch = c(16, 17),
  lty = 1,
  col = c("#0b6e69", "#a44a1f"),
  xlab = "Period",
  ylab = "Value",
  main = "Actuals vs forecast"
)
legend(
  "topright",
  legend = c("Actual", "Forecast"),
  col = c("#0b6e69", "#a44a1f"),
  lty = 1,
  pch = c(16, 17),
  bty = "n"
)
```

## Try it

1. Build a toy forecast-versus-actual dataset for two products or customers.
2. Compute the error summary.
3. Identify which series has the worst relative bias.

## Stretch

Use a monthly series such as `AirPassengers` or a small project series and ask:

1. what is the reporting level?
2. what is the measurement level?
3. what would a stakeholder want to see in an error report?

Then sketch the table you would want `himodel` to produce.

## Where it connects to other packages

- `hidate` for time boundaries
- `himunge` for shaping the input data
- `hipoint` for regime shifts and anomalies
- `higrate` if you want AI-generated narratives around modeled tables

## Think

What is the fastest way to lose stakeholder trust in a forecasting project:

- a small model miss
- or unclear, inconsistent error reporting?
