---
title: "Modeling and Forecasting"
---

# Where `himodel` fits

`himodel` is not the most universal package in the hiverse, but it is important
for forecasting and structured modeling work.

It is best understood as a modeling utility package with a strong forecasting
flavor, not as a general replacement for all of R's modeling ecosystem.

## What it is useful for

- standardized forecast error calculation
- simple univariate forecasting models
- prechecks that reject series not fit for a given model
- building your own univariate model objects
- forecast reshaping and reallocation

## `compute_error()`: standardize the evaluation layer

```{r}
library(himodel)

forecast_eval <- data.frame(
  sku = c("A", "A", "A", "B", "B", "B"),
  actual = c(100, 120, 110, 80, 75, 90),
  forecast = c(95, 130, 100, 82, 70, 88)
)

himodel::compute_error(
  forecast_eval,
  measure_level = c("sku"),
  report_level = c("sku"),
  actual_var = "actual",
  pred_var = "forecast"
)
```

The key idea is that `himodel` separates:

- the level at which forecasts were made
- the level at which results are reported

That is one reason it fits well in a shared forecasting workflow rather than as
just another collection of ad hoc helper functions.

## Built-in univariate models

`himodel` already ships a small family of simple univariate forecasting models.

```{r}
himodel::list_univariate_models()
```

These are not meant to replace every forecasting technique in R. They are
useful because they are simple, inspectable, and packaged in a common interface.

## Prechecks and a built-in model

The package has an explicit idea of a precheck: before you forecast anything,
decide whether the series is suitable for the method you want to run.

```{r}
library(dplyr)

all_dates <- seq(as.Date("2024-01-07"), by = "week", length.out = 64)

make_hist <- function(sku, offset, promo_bump) {
  actual_vals <- round(100 + offset + 8 * sin(seq_len(56) / 6) + seq_len(56) * 0.25, 1)
  future_actuals <- round(
    100 + offset + 8 * sin(57:64 / 6) + (57:64) * 0.25 +
      c(0, promo_bump, 0, 0, promo_bump, 0, 0, 0),
    1
  )

  list(
    actual = data.frame(
      sku = sku,
      date = all_dates[1:56],
      promo_strength = c(rep(0, 40), rep(0.2, 8), rep(0, 8)),
      actual = actual_vals
    ),
    holdout = data.frame(
      sku = sku,
      date = all_dates[57:64],
      actual = future_actuals
    )
  )
}

a <- make_hist("A", 0, 12)
b <- make_hist("B", -18, 9)

actual_df <- bind_rows(a$actual, b$actual)
future_df <- bind_rows(
  data.frame(sku = "A", date = all_dates[57:64], promo_strength = c(0, 0.3, 0, 0, 0.2, 0, 0, 0)),
  data.frame(sku = "B", date = all_dates[57:64], promo_strength = c(0, 0, 0.2, 0, 0, 0.3, 0, 0))
)
holdout_df <- bind_rows(a$holdout, b$holdout)
```

```{r}
lya_model <- himodel::make_model_lya("Weeks", generic_prechecks = TRUE)

himodel::precheck(
  lya_model,
  actual_df |>
    dplyr::filter(sku == "A") |>
    dplyr::select(date, actual, promo_strength),
  future_df |>
    dplyr::filter(sku == "A") |>
    dplyr::select(date, promo_strength)
)
```

That returns `TRUE` here, meaning the series is suitable for a last-year-average
model.

```{r}
lya_forecast <- himodel::run_univariate_model(
  lya_model,
  actual_df,
  future_df,
  group_vars = "sku"
)

lya_forecast
```

Now evaluate the forecast on a simple holdout.

```{r}
evaluation_df <- lya_forecast |>
  dplyr::left_join(holdout_df, by = c("sku", "date")) |>
  dplyr::transmute(sku, actual, forecast = total)

himodel::compute_error(
  evaluation_df,
  measure_level = "sku",
  report_level = "sku",
  actual_var = "actual",
  pred_var = "forecast"
)
```

## Why this matters

A lot of poor forecasting practice starts before the model itself:

- wrong aggregation level
- unfit time series sent into the wrong model
- poor error reporting
- hidden bias
- weak hierarchy handling

`himodel` is useful when you want a standard way to talk about those things.

## A simple visual

```{r}
plot_df <- actual_df |>
  dplyr::filter(sku == "A") |>
  dplyr::select(date, actual) |>
  dplyr::bind_rows(
    lya_forecast |>
      dplyr::filter(sku == "A") |>
      dplyr::transmute(date, actual = total)
  )

hist_df <- actual_df |>
  dplyr::filter(sku == "A") |>
  dplyr::select(date, actual)

forecast_df <- lya_forecast |>
  dplyr::filter(sku == "A") |>
  dplyr::transmute(date, forecast = total)

holdout_plot_df <- holdout_df |>
  dplyr::filter(sku == "A")

matplot(
  x = c(hist_df$date, forecast_df$date),
  y = cbind(
    c(hist_df$actual, rep(NA_real_, nrow(forecast_df))),
    c(rep(NA_real_, nrow(hist_df)), forecast_df$forecast),
    c(rep(NA_real_, nrow(hist_df)), holdout_plot_df$actual)
  ),
  type = "l",
  lty = 1,
  lwd = c(2, 2, 2),
  col = c("#0b6e69", "#a44a1f", "#3b3b3b"),
  xlab = "Week",
  ylab = "Value",
  main = "Built-in himodel forecast vs holdout actuals"
)
legend(
  "topright",
  legend = c("History", "Forecast", "Holdout actual"),
  col = c("#0b6e69", "#a44a1f", "#3b3b3b"),
  lty = 1,
  bty = "n"
)
```

## Define your own univariate model

One of the more underrated ideas in `himodel` is that a model can be a small
object plus S3 methods. If you can define:

- `precheck.<class>()`
- `estimate.<class>()`
- `model_version.<class>()`

then you can plug your own model into the same workflow.

```{r}
make_model_demo_mean <- function(freq = "Weeks", lookback = 8, generic_prechecks = TRUE) {
  model <- list(
    freq = freq,
    lookback = lookback,
    generic_prechecks = generic_prechecks
  )
  class(model) <- "demo_mean"
  model$version <- himodel::model_version(model)
  model
}

precheck.demo_mean <- function(method, actual_df, future_df, ...) {
  if (method$generic_prechecks) {
    himodel::generic_prechecks(method$freq, actual_df, future_df)
  }
  nrow(actual_df) >= method$lookback
}

estimate.demo_mean <- function(method, actual_df, future_df, ...) {
  avg <- mean(utils::tail(actual_df$actual, method$lookback))
  dplyr::mutate(future_df, baseline = avg, total = avg)
}

model_version.demo_mean <- function(method, ...) {
  "0.1.0"
}

demo_model <- make_model_demo_mean()
demo_model$version

himodel::run_univariate_model(
  demo_model,
  actual_df,
  future_df,
  group_vars = "sku"
)
```

## Try it

1. Run `list_univariate_models()` and pick one model you have not used before.
2. Build a small weekly series and run `precheck()` on it.
3. Forecast a short holdout and compute the error summary.
4. Decide whether the model is plausible or just a baseline.

## Stretch

Create a small custom model object that does something simple but explicit, for
example:

1. rolling mean of the last `k` periods
2. last non-zero value
3. mean of the last year, but only on non-promo periods

Then add:

1. a `precheck.<class>()`
2. an `estimate.<class>()`
3. a `model_version.<class>()`

and run it with `run_univariate_model()`.

## Where it connects to other packages

- `hidate` for time boundaries
- `himunge` for shaping the input data
- `hipoint` for regime shifts and anomalies
- `higrate` if you want AI-generated narratives around modeled tables

## Think

What is the fastest way to lose stakeholder trust in a forecasting project:

- a slightly weak model
- or a workflow that cannot explain why the model should be trusted?
